Chemical tagging: \\ Identification of stellar phase-space structures \\ purely by chemical-abundance similarity

### 2016-01-15 papers/clusters.tex

% to-do
% - write data and cannon section
% - write method and results section
% - write discussion section
% - put in references
% - cite codes!
% - put in acknowledgement
% - fix all remaining occurrences of ARC, MKN, HWR, or DWH
% - forward to Freeman, others?
% - submit to journal and arXiv

@@

% text shit
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\aspcap}{\project{\acronym{ASPCAP}}}
\newcommand{\gaiaeso}{\project{Gaia--\acronym{ESO}}}
\newcommand{\galah}{\project{\acronym{GALAH}}}
\newcommand{\thecannon}{\project{The~Cannon}}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\eg}{\foreign{e.\,g.}}
\newcommand{\etal}{\foreign{et\,al.}}
\newcommand{\documentname}{\textsl{Article}}

@@

If every star in the Milky Way was born in a molecular cloud that has
unique chemical-abundance characteristics (as a function of time), and
if every star preserves its surface abundances (or significant aspects
of them) over its lifetime, then stars with common birthplaces and
time ought to be identifiable in their surface abundances.
That is, it ought to be possible to use stellar abundances to link up
co-eval stars---even if at the present day they are no longer close to
one another in phase space---and infer a detailed star-formation and
dynamical history for the Galaxy.
This idea has been dubbed ``chemical tagging''.
It was proposed long ago \citep{Freeman2002} and is one of the
principal motivations for a number of surveys, most notably
\apogee\ \citep{apogee}, \gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah13, galah15}.
In order to measure chemical tags and infer the Galaxy's dynamical
properties and formation history, these surveys are each measuring high resolution,
high signal-to-noise spectra for hundreds of thousands of stars across
the Galaxy's disk, bulge and halo.

Chemical tagging holds the promise of revealing not just the
star-formation history of the Galaxy, but also the accretion history
(as things that fall in are expected to be chemically distinct from
those that form in the parent body; \citealt{something}) and
stellar-orbit diffusion process like radial mixing and radial
migration (for example, \citealt{Roskar2009, Quillen2015}).  After
stars are born---or after a star cluster is accreted and
disrupted---associations or groups will disperse, through either
coherent orbital evolution or else incoherent diffusion, but the
chemical-abundance tags can in principle make the former members
identifiable as co-eval stellar siblings.

@@

The second issue for chemical tagging---and the one we address
here---is measurement precision (and accuracy).
The current precision on abundance measurements is not high enough
(for example, \citealt{Ting2015, Martel2015}).
That said, it is clear that the \emph{data} are precise enough;
there is enough signal-to-noise at the relevant locations in spectrum
space to deliver high-precision tags.
Our view is that the traditional methods for making abundance
measurements just aren't mature enough:
There are issues with the physical models, and the data-analysis
methods do not make use of all of the information in the data sets.
Improving the models and exploiting the entire information
content in the data is critical if we are going to deliver useful
chemical tags.

Our contribution in this space has been to develop
\thecannon\ \citep{cannon, ages}, which is a data-driven model for
stellar spectra.
This model can deliver stellar parameters and chemical abundances to
stars, making use of every pixel of every stellar spectrum (that is,
all the information in the data) but making no use at all of physical
models of stars.
It relies only on there being training data---\emph{some} stars for
which parameters and abundances are known and believed.
In companion papers \citep{casey16, ness16} we show that
\thecannon\ can deliver 15 to 19 abundances for stars in the
\apogee\ survey, at precisions higher than even the training data on
which the model is trained.
We say more about this below.
We will show here that \thecannon\ improves chemical abundance measurements
to the point that \emph{chemical tagging is now possible.}
We have thereby effectively saved the collective asses of all the
collaborators in all of the above-referenced large spectroscopic
surveys.

@@

In what follows, we are going to use \apogee\ \acronym{DR12} data, in
which we can measure 15 element abundances (C, N, O, Na, Mg, Al, Si,
S, K, Ca, Ti, V, Mn, Fe, Ni).
The \apogee\ data set covers a huge radial extent and---because the
data are taken in the infrared---explores the thin disk.
However, it has the disadvantage that it's spatial coverage is swiss
cheese, which makes it hard to see, within the data set, linear or
extended stellar structures.
In many ways, \galah\ will deliver improvements, both because it will
have more abundances (possibly 29) and contiguous sky coverage.

@@

The details about how we select, reduce, and analyze the \apogee\ data
are given in full detail in the companion papers \citep{casey16, ness16}.
We summarize briefly here.
For the spectroscopic data, we use the \apogee\ \acronym{DR12}
(\cite{dr12}) spectroscopic data reductions for red-giant stars
($\logg < 3.9$) that are not flagged as bad in the
\apogee\ \aspcap\ (\cite{aspcap}) pipeline reductions.
There are XXXXX stars that make our $\logg$ and flag cuts.

@@

For the training data labels (used to train \thecannon; see below), we
use the \apogee\ \acronym{DR12} \aspcap\ atmospheric parameters and
15-element abundances, but only for the subset of XXXXX red-giant
stars that have spectral signal-to-noise ratio measurements greater
than 200.
This is a small fraction of the total sample, but these high
signal-to-noise stars are observed to obtain consistent, low-scatter,
sensible abundance-space measurements.

@@

Although only hints of small-scale clustering in the abundance space
are visible in \figurename~\ref{fig:all}, exploration of the data indicates that
known clusters do appear in the abundance space as over-densities.
This encourages us to look for over-densities automatically in the
abundance space and see if anything found that way would be over-dense
in phase space.
The simplest method for clustering points in $D$-dimensional space is
the \emph{k-means} algorithm (DWH CITE).
Briefly, the k-means algorithm is to find the points closest (in
$D$-space) to each of $K$ centers, and then take the mean of those
points to update the locations of the centers; iterate to convergence.
The output is the locations of the $K$ centers and the assignments of
all points to centers.
This algorithm is fast and performs well in practice in problems of
this nature.

The k-means algorithm has disadvantages, of course.
One is that $K$ must be chosen by hand (or heuristically at best).
Here we are only demonstrating a concept; we don't need to have the
best possible clustering.
For this reason we simply choose $K=128, 256, 512$ and look at all the
results.
The algorithm also only performs local optimization.
At each $K$ we perform 32 restarts with different initializations, and
preserve the best clustering (best according to the k-means score).

@@

All of the code for the clustering and visualization is built on numpy
(DWH CITE), scikit-learn (DWH CITE), and matplotlib (DWH CITE).

\begin{figure}[!p]
\insanefigure{cluster_0256_0253}
\caption{Whatevsinator.\label{fig:infallcluster}}
\end{figure}

@@

\acknowledgements
DWH: It is a pleasure to thank...

DWH: grants and grant numbers.

DWH: SDSS-III and SDSS-IV.

DWH: NASA ADS

DWH: github repo and hash.

DWH: Be sure to cite software a la the new ApJ guidelines.

\clearpage
\begin{thebibliography}{24}\raggedright
\bibitem[Ness \etal(2016)]{ness16}
whatev
\end{thebibliography}

>>>>

% to-do
% - write data and cannon section
% - write method and results section
% - write discussion section
% - put in references
% - cite codes!
% - fix all remaining occurrences of ARC, MKN, HWR, DWH, or CITE
% - forward to Freeman, others?
% - submit to journal and arXiv

@@

% text shit
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\sdssiii}{\project{\acronym{SDSS-III}}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\aspcap}{\project{\acronym{ASPCAP}}}
\newcommand{\gaiaeso}{\project{Gaia--\acronym{ESO}}}
\newcommand{\galah}{\project{\acronym{GALAH}}}
\newcommand{\thecannon}{\project{The~Cannon}}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\eg}{\foreign{e.\,g.}}
\newcommand{\etal}{\foreign{et\,al.}}
\newcommand{\documentname}{\textsl{Article}}

@@

If every star in the Milky Way was born in a molecular cloud that has
unique chemical-abundance characteristics (as a function of time), and
if every star preserves its surface abundances (or significant aspects
of them) over its lifetime, then stars with common birthplaces and
time ought to be identifiable in their surface abundances.
That is, it ought to be possible to use stellar abundances to link up
co-eval stars---even if at the present day they are no longer close to
one another in phase space---and infer a detailed star-formation and
dynamical history for the Galaxy.
This idea has been dubbed ``chemical tagging''.
It was proposed long ago \citep{freeman} and is one of the
principal motivations for a number of surveys, most notably
\apogee\ \citep{apogee}, \gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah}.
In order to measure chemical tags and infer the Galaxy's dynamical
properties and formation history, these surveys are each measuring high resolution,
high signal-to-noise spectra for hundreds of thousands of stars across
the Galaxy's disk, bulge and halo.

Chemical tagging holds the promise of revealing not just the
star-formation history of the Galaxy, but also the accretion history
(as things that fall in are expected to be chemically distinct from
those that form in the parent body; \citealt{something}) and
stellar-orbit diffusion process like radial mixing and radial
migration (for example, \citealt{Roskar2009, quillen}).  After
stars are born---or after a star cluster is accreted and
disrupted---associations or groups will disperse, through either
coherent orbital evolution or else incoherent diffusion, but the
chemical-abundance tags can in principle make the former members
identifiable as co-eval stellar siblings.

@@

The second issue for chemical tagging---and the one we address
here---is measurement precision (and accuracy).
The current precision on abundance measurements is not high enough
(for example, \citealt{Ting2015, martel}).
That said, it is clear that the \emph{data} are precise enough;
there is enough signal-to-noise at the relevant locations in spectrum
space to deliver high-precision tags.
Our view is that the traditional methods for making abundance
measurements just aren't mature enough:
There are issues with the physical models, and the data-analysis
methods do not make use of all of the information in the data sets.
Improving the models and exploiting the entire information
content in the data is critical if we are going to deliver useful
chemical tags.

Our contribution in this space has been to develop
\thecannon\ \citep{thecannon, ages}, which is a data-driven model for
stellar spectra.
This model can deliver stellar parameters and chemical abundances to
stars, making use of every pixel of every stellar spectrum (that is,
all the information in the data) but making no use at all of physical
models of stars.
It relies only on there being training data---\emph{some} stars for
which parameters and abundances are known and believed.
In companion papers \citep{casey16, ness16} we show that
\thecannon\ can deliver 15 to 19 abundances for stars in the
\apogee\ survey, at precisions higher than even the training data on
which the model is trained.
We say more about this below.
We will show here that \thecannon\ improves chemical abundance measurements
to the point that \emph{chemical tagging is now possible.}
We have thereby effectively saved the collective asses of all the
collaborators in all of the above-referenced large spectroscopic
surveys.

@@

In what follows, we are going to use \apogee\ \acronym{DR12} data \citep{dr12}, in
which we can measure 15 element abundances (C, N, O, Na, Mg, Al, Si,
S, K, Ca, Ti, V, Mn, Fe, Ni).
The \apogee\ data set covers a huge radial extent and---because the
data are taken in the infrared---explores the thin disk.
However, it has the disadvantage that it's spatial coverage is swiss
cheese, which makes it hard to see, within the data set, linear or
extended stellar structures.
In many ways, \galah\ will deliver improvements, both because it will
have more abundances (possibly 29) and contiguous sky coverage.

@@

The details about how we select, reduce, and analyze the \apogee\ data
are given in full detail in the companion papers \citep{casey16, ness16}.
We summarize briefly here.
For the spectroscopic data, we use the \apogee\ \acronym{DR12}
(\cite{dr12}) spectroscopic data reductions for red-giant stars
($\logg < 3.9$) that are not flagged as bad in the
\apogee\ \aspcap\ (\cite{aspcap}) pipeline reductions.
There are 98462 [DWH CHECK] stars that make our $\logg$ and flag cuts.

@@

For the training data labels (used to train \thecannon; see below), we
use the \apogee\ \acronym{DR12} \aspcap\ atmospheric parameters and
15-element abundances, but only for the subset of ARC:XXXXX red-giant
stars that have spectral signal-to-noise ratio measurements greater
than 200.
This is a small fraction of the total sample, but these high
signal-to-noise stars are observed to obtain consistent, low-scatter,
sensible abundance-space measurements.

@@

Although only hints of small-scale clustering in the abundance space
are visible in \figurename~\ref{fig:all}, exploration of the data indicates that
known clusters do appear in the abundance space as over-densities.
This encourages us to look for over-densities automatically in the
abundance space and see if anything found that way would be over-dense
in phase space.
The simplest method for clustering points in $D$-dimensional space is
the \emph{k-means} algorithm (see CITE for a pedagogical introduction).
Briefly, the k-means algorithm is to find the points closest (in
$D$-space) to each of $K$ centers, and then take the mean of those
points to update the locations of the centers; iterate to convergence.
The output is the locations of the $K$ centers and the assignments of
all points to centers.
This algorithm is fast and performs well in practice in problems of
this nature.

The k-means algorithm has disadvantages, of course.
One is that $K$ must be chosen by hand (or heuristically at best).
Here we are only demonstrating a concept; we don't need to have the
best possible clustering.
For this reason we simply choose $K=128$, $K=256$, and $K=512$ and look at all the
results.
The algorithm also only performs local optimization.
At each $K$ we perform 32 restarts with different initializations, and
preserve the best clustering (best according to the k-means score).

@@

All of the code for the data manipulation, clustering, and
visualization is built on numpy (CITE), scikit-learn (CITE),
and matplotlib (CITE).

\begin{figure}[!p]
\insanefigure{cluster_0256_0253}
\caption{Whatevsinator.\label{fig:M13}}
\end{figure}

@@

\acknowledgements
DWH was partially supported by
  the NSF (grants IIS-1124794, AST-1517237),
  NASA (grant NNX12AI50G), and
  the Moore-Sloan Data Science Environment at NYU.
This research made use of the NASA \project{Astrophysics Data System}.

This project made use of \sdssiii\ data.
Funding for \sdssiii\ has been provided by the Alfred P. Sloan
Foundation, the Participating Institutions, the National Science
Foundation, and the U.S. Department of Energy Office of Science. The
\sdssiii\ web site is http://www.sdss3.org/.

\sdssiii\ is managed by the Astrophysical Research Consortium for the
Participating Institutions of the \sdssiii\ Collaboration including the
University of Arizona, the Brazilian Participation Group, Brookhaven
National Laboratory, Carnegie Mellon University, University of
Florida, the French Participation Group, the German Participation
Group, Harvard University, the Instituto de Astrofisica de Canarias,
the Michigan State/Notre Dame/JINA Participation Group, Johns Hopkins
University, Lawrence Berkeley National Laboratory, Max Planck
Institute for Astrophysics, Max Planck Institute for Extraterrestrial
Physics, New Mexico State University, New York University, Ohio State
University, Pennsylvania State University, University of Portsmouth,
Princeton University, the Spanish Participation Group, University of
Tokyo, University of Utah, Vanderbilt University, University of
Virginia, University of Washington, and Yale University.

\clearpage
\begin{thebibliography}{24}\raggedright
\bibitem[Alam \etal(2015)]{dr12}
Alam,~S., Albareti,~F.~D., Allende~Prieto,~C., \etal, 2015, \apjs, 219, 12
\bibitem[Blanco-Cuaresma \etal(2015)]{blancocuaresma}
Blanco-Cuaresma,~S., Soubiran,~C., Heiter,~U., \etal, 2015, \aap, 577, A47
\bibitem[Bovy(2015)]{bovy}
Bovy,~J., 2015, arXiv:1510.06745
\bibitem[Casey \etal(2016)]{casey16}
Casey,~A.~R., Hogg,~D.~W., Ness,~M., Rix,~H.-W., 2016, in preparation
\bibitem[De~Silva \etal(2015)]{galah}
De~Silva,~G.~M., Freeman,~K.~C., Bland-Hawthorn,~J., \etal, 2015, \mnras, 449, 2604
\bibitem[Freeman \& Bland-Hawthorn(2002)]{freeman}
Freeman,~K., \& Bland-Hawthorn,~J., 2002, \araa, 40, 487
\bibitem[Garc{\'{\i}}a~P{\'e}rez \etal(2015)]{aspcap}
Garc{\'{\i}}a~P{\'e}rez,~A.~E., Allende~Prieto,~C., Holtzman,~J.~A., \etal, 2015, arXiv:1510.07635
\bibitem[Majewski \etal(2015)]{apogee}
Majewski,~S.~R., Schiavon,~R.~P., Frinchaboy,~P.~M., \etal, 2015, arXiv:1509.05420
\bibitem[Melendez \etal(2014)]{melendez}
Mel{\'e}ndez,~J., Ram{\'{\i}}rez,~I., Karakas,~A.~I., \etal, 2014, \apj, 791, 14
\bibitem[Mitschang \etal(2014)]{mitschang}
Mitschang,~A.~W., De~Silva,~G., Zucker,~D.~B., \etal, 2014, \mnras, 438, 2753
\bibitem[Ness \etal(2015)]{thecannon}
Ness,~M., Hogg,~D.~W., Rix,~H.-W., Ho,~A.~Y.~Q., Zasowski,~G., 2015, \apj, 808, 16
\bibitem[Ness \etal(2016a)]{ages}
Ness,~M., Hogg,~D.~W., Rix,~H.-W., \etal, 2016a, \apj, in press (arXiv:1511.08204)
\bibitem[Ness \etal(2016b)]{ness16}
Ness,~M., Hogg,~D.~W., Casey,~A.~R., Rix,~H.-W., 2016b, in preparation
\bibitem[Quillen \etal(2015)]{quillen}
Quillen,~A.~C., Anguiano,~B., De~Silva,~G., \etal, 2015, \mnras, 450, 2354
\bibitem[Ting \etal(2016)]{ting}
Ting,~Y.-S., Conroy,~C., \& Rix,~H.-W., 2016, \apj, in press (arXiv:1507.07563)
\end{thebibliography}
